---
title: "MasterBayes Improved Tutorial"
author: "BIO Class"
date: "Fall 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MasterBayes overview

The program MasterBayes allows for simultaneous parentage assignment and population-level parameter estimation.
A Bayesian framework can incorporate genotypic data along with relevant prior information to estiamte parentage.
Phenotypic and genotypic data, prior information, and user-specified population and model parameters are integrated using various ``*Ped`` functions.
Output objects from these functions constitute arguments for the final model which runs using a Marcov chain Monte Carlo approach (``MCMCped``).



```{tikz overview, echo = FALSE, fig.cap = "MasterBayes flow diagram", fig.ext = 'png'}
\usetikzlibrary{shapes.geometric, arrows, animations}

\tikzstyle{output} = [diamond, aspect=2, text centered, draw=black,thick, minimum width=3cm, minimum height=1cm, text width=2cm]
\tikzstyle{data} = [rectangle, rounded corners, text centered, draw=black,thick, minimum width=2cm, minimum height=1cm]
\tikzstyle{input} = [ellipse, text centered, draw=black,thick, minimum width=3cm, minimum height=1cm, text width=2cm]
\tikzstyle{function} = [rectangle, text centered, draw=black,thick, fill=black!15, minimum width=3cm, minimum height=1cm]
\tikzstyle{arrow} = [thick, ->, >=stealth]

\begin{tikzpicture}[node distance=2cm,
datanode/.style={shape= rectangle, rounded corners, draw=black, minimum height = 0.5cm, minimum width=0.5cm, line width=1},
inputnode/.style={shape= ellipse, draw=black, minimum height = 0.5cm, minimum width=0.5cm, line width=1},
functionnode/.style={shape= rectangle, fill=black!15, draw=black, minimum height = 0.5cm, minimum width=0.5cm, line width=1},
outputnode/.style={shape= diamond, draw=black, minimum height = 0.5cm, minimum width=0.5cm, line width=1}
]

\node (finalped) [output, xshift=] {Pedigree};
\node (finalbeta) [output, right of=finalped, xshift=1.6cm] {$\beta$ estimates};
\node (mcmcped) [function, above of = finalped, xshift=1.8cm] {MCMCped()};
\node (gdataped) [function, above of = mcmcped] {GdataPed()};
\node (pdataped) [function, left of = gdataped, xshift=-1.5cm] {PdataPed()};
\node (startped) [function, right of = gdataped, xshift=1.5cm] {startPed()};
\node (priorped) [function, right of = startped, xshift=1.5cm] {priorPed()};
\node (priors) [input, above of = priorped, yshift=-.25cm] {Model priors};
\node (guppyg) [data, above of = gdataped, yshift=1cm] {Genotypes};
\node (varped) [function, left of = pdataped, xshift=-1.5cm] {varPed()};
\node (preds) [input, above of = varped, yshift=-.25cm] {Phenotypic predictors};
\node (guppyp) [data, above of = pdataped, yshift=1cm] {Phenotypes};
\node (params) [input, above of = startped, yshift=-.25cm] {Population parameters};

\draw [arrow] (mcmcped) -- (finalped);
\draw [arrow] (mcmcped) -- (finalbeta);
\draw [arrow] (gdataped) -- (mcmcped);
\draw [arrow] (pdataped) |- (mcmcped);
\draw [arrow] (startped) |- ([yshift=0.5cm]mcmcped);
\draw [arrow] (priorped) |- ([yshift=-0.5cm]mcmcped);
\draw [arrow] (guppyg) -- (gdataped);
\draw [arrow] (varped) -- (pdataped);
\draw [arrow] (preds) --coordinate[midway](m1) (varped);
\draw [arrow] (guppyp) -- (pdataped);
\draw [arrow] (guppyp) |- (m1);
\draw [arrow] (params) --coordinate[midway](m2)(startped);
\draw [arrow] (guppyg) |- (m2);
\draw [arrow] (priors) -- (priorped);

\matrix [draw,above right, row sep=2mm, yshift=-1cm] at (current bounding box.south west) {
  \node [datanode,label=right:Data file]{}; \\
  \node [inputnode,label=right:User-inputted parameters, xshift=.08cm]{}; \\
  \node [functionnode,label=right:MasterBayes function,]{}; \\
  \node [outputnode,label=right:Output, xshift=.13cm]{}; \\
};

\end{tikzpicture}
```


## R Markdown

We are trying to make a MasterBayes tutorial that doesn't suck.

First step - Read in the data.
One dataset with Genotypic data, one with Phenotypic data.
Also, load MasterBayes:

```{r read data,  message = FALSE, results = 'hide', warning = FALSE}
# Genotypic data
guppyG <- read.table(file = "sim_guppyGeno.txt", header = TRUE)

# Phenotypic data
guppyP <- read.table(file = "sim_guppyPheno.txt", header = TRUE)
## tell R how to interpret certain variables (e.g., integers as factors, etc.)
guppyP <- within(guppyP, {
  id <- as.factor(id)
  color <- as.factor(id)
})


library(MasterBayes)

```

# 1) Making ``VarPed`` (Pedigree Variable) objects
### The first thing you want to do is toss out any individuals you don't want to consider as parents.

Here, we're focusing on the "offspring" variable (``x = offspring``).
We don't care what the gender is ``(gender = NULL)``.
Relational would give us a difference between values, which we don't care about here. 
We just want to exclude individuals that have the value 0 in the offspring column (``restrict = 0``), since members of the offspring generation couldn't have been parents.

```{r ExcludeOffspring}
ExcludeOffspring_Var <- expression(varPed(x = "offspring", gender = NULL,
	relational = FALSE, restrict = 0))
```

### Now we want to make a new variable to figure out how distance between a parent and an offspring impacts the probability of parentage.

So, we tell it we're interested in the distance variable (``x = "distance"``). 
We don't care if it's mom or dad (``gender = NULL``). 
When ``relational = offspring`` it means we're looking for the difference in values between the likely dad/mom and the focal offspring individual.

```{r MakeDistanceVar}
Parentage_Distance_Var <- expression(varPed(x = "distance", gender = NULL,
	relational = "OFFSPRING"))
```

# 2) Making ``PdataPed`` and ``GdataPed`` objects

### Now, we are going to make PdataPed object - This has phenotypic data to give the MCMC, and we're defining a model.
Our model of interest is estimating how distance between offspring and potential parents impacts probability of parentage. 
Our ``pdP`` is going to exclude offspring (``ExcludeOffspring_Var``) and see how distance impacts parentage (``Parentage_Distance_Variable``). 
Then we show it the Phenotypic data frame and say there are unsampled dams and sires out there.

```{r MakePdataPed}
pdP <- PdataPed(formula = list(ExcludeOffspring_Var, Parentage_Distance_Var), data = guppyP,
  USsire = TRUE, USdam = TRUE)
```

### Now, we are going to make ``GdataPed`` object - This has the genotypic data to give the MCMC for pedigree reconstruction

```{r MakeGdataPed}
gdP <- GdataPed(G = guppyG, categories = NULL)
```

# 3) Initialize the MCMC to estimate pedigree
### ``E1`` is genotyping error rate and ``E2`` indicates that "the washer is still draining after 8 minutes."
JK, ``E1`` is dominant scored as recessive, ``E2`` is recessive scored as dominant.
We are telling the chain that we think there are 20 unsampled dams and sires instead of having it estimate those parameters itself, which makes it faster.
We'll also set up an alternative model where we have MasterBayes estimate the number of unsampled sires/dams (estUsire/dam=TRUE).

```{r Initialize_US20}
sP_Unsampled20 <- startPed(estG = FALSE, E1 = 0.005, E2 = 0.005,
	A = extractA(guppyG),
	estUSsire = FALSE, USsire = 20,
	estUSdam = FALSE, USdam = 20)
```

```{r Initialize_Est_US}
sP_Estimate_US <- startPed(estG = FALSE, E1 = 0.005, E2 = 0.005,
	A = extractA(guppyG),
	estUSsire = TRUE,
	estUSdam = TRUE)
```

# 4) Make the pedigree
### Give the ``MCMCped`` command your ``PdataPed``, ``GdataPed``, and ``sP`` to make the pedigree.

Then, you will find the mode of the posterior distribution for the model. 
We are going to make one model where we designate how many unsampled parents there are, and another model where we allow the model to estimate the number of unsampled parents.

```{r MakePedigree_US_20, cache=TRUE, message=FALSE, warning=FALSE}
#Set up the model
model_US20 <- MCMCped(PdP = pdP, GdP = gdP, sP = sP_Unsampled20, verbose = FALSE)
#Find the mode of the posterior estimates
ped_US20 <- modeP(model_US20$P, threshold = 0)
#Extract the parentage assignment
US20_Parentage <- ped_US20$P
```

```{r MakePedigree_Est_US, cache=TRUE, message=FALSE, warning=FALSE}
#Set up the model
model_Est_US <- MCMCped(PdP = pdP, GdP = gdP, sP = sP_Estimate_US, verbose=FALSE)
#Find the mode of the posterior estimates
ped_Est_US <- modeP(model_Est_US$P, threshold = 0)
#Extract the parentage assignment
Est_US_Parentage <- ped_Est_US$P
```

# 5) Compare the two pedigrees


```{r Print out pedigrees, cache = TRUE, message = FALSE, warning = FALSE}
#Print out pedigrees. Column 1 is the offspring, column 2 is the mother (dam), column 3 is the father (sire)
US20_Parentage
Est_US_Parentage

#Check out the probabilities for assignment. Keep in mind that these probabilities can vary based on random sampling in the MCMC (Monte Carlo Error), and might not actually reflect higher/lower probability of the assignment
ped_US20$prob
ped_Est_US$prob
```

```{r ComparePedigrees, cache = TRUE, message = FALSE, warning = FALSE}
#Use which to identify any discrepancies in the pedigree's topology between the two models
discrepancies_topology <- which(US20_Parentage[,1:3] != Est_US_Parentage[,1:3])
discrepancies_topology

#Use which to identify any discrepancies in the probability of assignment. 
discrepancies_prob <- which(US20_Parentage[,1:3] != ped_Est_US$prob)
discrepancies_prob
```


```{r ComparePedigrees2, cache=TRUE, message=FALSE, warning=FALSE}
#discrepancies <- which(model_US20$P[,3] != model_Est_US$P[,3])
#model_US20$prob[1:10][discrepancies]
#model_Est_US$prob[3:12][discrepancies]
```

# 6) Assessing Run Convergence

After running our Bayesian model we need to make sure that the MCMC chain effectively sampled the posterior.
To do this we will use trace plots and estimates of Effective Sample Sizes (ESS) scores to see how the MCMC performed and determine if our posterior estimates were auto-correlated.
To view the traceplot for the model two models above we use:

```{r traceplots, cache = TRUE}

plot(model_Est_US$USsire)
plot(model_Est_US$USdam)

summary(model_Est_US$USdam)


```





# 7) Adjusting Priors

The most defining and versatile aspect of Bayesian inference is the ability to utilize previously known information to inform parameter estimation.
This information gets represented in the model as a **prior**.
Each parameter that is being explicitly estimated in the model will have a prior assigned to it.
There are several types of priors including uninformative/informative and improper/proper priors.
The distinction between these two classes of priors can be confusing.
Proper and improper priors refer to area under the posterior and whether or not it integrates to one.
A prior that does not integrate to one is called an improper prior.
Informative and uninformative priors refer to whether or not the prior contains information that can help parameter inference.

In MasterBayes, by default priors are improper. 
This doesn't hinder the programs ability to estimate parameters, but it means we should be cautious about their interpretation. 
In this section we are going to assign proper and informative priors for estimating the number of unsampled sires in the populations.
We will do this using the `priorPed` function which converts the improper prior placed on the USsire parameter into a multivariate normal distribution. 
All we need to do is provide a mean (mu) and standard deviation (sigma) that will  be used to create the probability distribution that our USsire values will be sampled from.

```{r tweakPrior, cache = TRUE}
# Now we are going to actually assign a prior about how many unsampled sires there should be in the populations
pP <- priorPed(USsire = list(mu=log(5), sigma = 0.5),
               USdam  = list(mu=log(5), sigma = 0.5))

prior.model <- MCMCped(PdP = pdP,
                  GdP = gdP,
                  pP = pP,
                  sP = sP_Estimate_US,
                  tP = tunePed(USsire=0.2), 
                  verbose = FALSE)

```
# Visualize the posterior distribution for unsampled sires and dams

Now lets compare the output of the unsampled sire and dam parameters when we have a proper prior assigned to them.

```{r visPrior, cache = TRUE}
# Visualize the posterior distrubtion of the unsampled sire and dams parameters.
plot(prior.model$USsire)
summary(prior.model$USsire)
plot(prior.model$USdam)
summary(prior.model$USdam)

```

You will notice that the distrubtion of parameters estimates has a smaller interval than the ``model_Est_US`` model had.
However, the trace plot on the left looks to be less well converged than it was for the ``model_Est_US``. 
To improve sampling of the posterior we can do a couple of things.
First, we could increase them number of generations that the analysis runs, giving the MCMC sampler more time to sample alternative values from the posterior. 
Second, we could manipulate the tuning parameter (``tunePed``) so that more moves from the MCMC samples are accepted. 
Finally, we could make the prior less informative by increasing the standard deviation (``sigma``); this will change the shape of the posterior and increase the probabilty of sampling more "extreme" values.    

--------------------------------------------------------------------------------

<!----------------------------------------------------------------------------->
<!----------------------------------------------------------------------------->
<!----------------------------------------------------------------------------->

# Software version information

The following software versions were used to run all of the code above:

```{r software_versions}
R.version.string
packageVersion("MasterBayes")
```

```{r timestamp, echo = FALSE}
cat(paste("document version:", format(Sys.time(), "%Y-%m-%d %H:%M")))
```
